Music Recommendation System Using the Million Song Dataset

### 2015-05-08 writeup/MSDgroupwriteup.tex
Update MSDgroupwriteup.tex

\begin{abstract}
A recommendation system was attempted to recommend songs to users using data from the Million Song Dataset available on Kaggle. The main focus of this paper was using content-based collaborative filtering algorithms that only utilize user feedback in order to predict what songs users may like. For benchmarking the algorithms, the same Mean Average Precision truncated at 500 recommended songs used for the original Kaggle competition was utilized. It was discovered that probabilistic matrix factorization did not improve results much from using a baseline of simply recommending popular songs with MAP values around 0.014, while artist-based popularity along with user-based and item-based collaborative filtering methods yielded much better results exceeding MAP of 0.04.

@@

\section{Introduction}
In the online world, recommendation systems are becoming increasingly important in various industries including retail, entertainment, and even dating. 
Because of their popularity and usefulness, there are numerous incentives to implement a good recommendation system. 
In the music industry, companies use recommendation systems to provide a better service to their listeners. 
With the music industry shifting away from a more traditional distribution model of physical CDs or records, 
having an efficient and well performing recommendation system is very important.

In building a recommendation system, we had to take into account various trade-offs in the type of models we chose to fit. For example, some methods take more information about the user and the song into account (i.e. content-based) while others use a collaborative filtering approach without taking into account user or song characteristics such as age, genre, location, etc. In this paper, we focus on collaborative filtering methods.
There is no set similarity measure between two songs, as evident by the fact that songs by the same artist are often very different. As such, there are many different types of algorithms to explore.

For our project, we explored various song recommendation systems algorithms such as popularity recommendations, artist based recommendations, matrix factorization, and user and item based similarity measures.

@@

\subsection{Testing}
We tested our algorithms against each other using a mean average precision 
score for each method.  The mean average precision (MAP) calculation came directly from the Kaggle MSD challenge \citep{McFee:2012:MSD:2187980.2188222}. The score was used to rank the competition entries.

Specifically, we used MAP@500 scores which take into account the first 500 recommendations given to each user. Generally, the score looks at the precision between a list of recommendations for a user and that user's test set. This was done by calculating the average precision at each element in the list of recommended items (i.e. percentage of correct items among first $k$ recommendations) and averaging these for the first 500 recommendations (i.e. for each $k \in \{1, ..., 500\}$, calculate the precision and average them together).
We then averaged the score over all users. Intuitively, the score looks at the percentage of recommendations that were in-line with the test set of songs actually listened to but also takes order into account. So it is preferable to recommend the songs in the user's test set earlier in the list of 500. The MAP score is similar to the AUC of the ROC curve discussed in class except it takes ordering of the recommendations into account, which is an important factor to consider in a recommendation system.

We used two different baseline benchmarks \citep{McFee:2012:MSD:2187980.2188222} to compare our MAP scores against. The first was to simply recommend the top 500 most popular songs to every user. This naive method resulted in a MAP score of 0.0138 (top 500 songs based on number of users who listened to the song) or 0.0126 (top 500 songs based on number of plays the song had). The second method was to first recommend a user the most popular songs of artists the user already had already played (but different songs from that artist). If this did not result in enough recommendations, we would then recommend the overall most popular songs as well at the end (similar to baseline method 1). Although this method is conservative in that it does not go beyond the tastes the user already has in the training set, it resulted in a significant improvement in the MAP score to 0.0448.

>>>>

\begin{abstract}
We attempted to build a recommendation system to recommend songs to users using data from the Million Song Dataset available on Kaggle. The main focus of this paper was to use collaborative filtering algorithms that only utilize user feedback in order to predict what songs users may like. For benchmarking the algorithms, the same Mean Average Precision truncated at 500 recommended songs used for the original Kaggle competition was chosen. It was discovered that probablistic matrix factorization did not improve results much from using a baseline of simply recommending popular songs with MAP values around 0.014, while artist-based popularity along with user-based and item-based collaborative filtering methods yielded much better results exceeding a MAP value of 0.04.

@@

\section{Introduction}
With an increasing amount of accessible data, recommendation systems are becoming increasingly viable and important in various industries including retail, entertainment, and even dating. Recommendation systems allow for services to extend their reach to customers by allowing their users to discover things of interest. Hence, there is a strong interest in implementing a good recommendation system so that their user base can grow. It can be said that building such a system is even more important for the music industry because the entire library of artists and songs are far larger compared to something like the movie industry so recommending the right music to a specific user is an even harder task. On the other hand, with the shift from a traditional distribution model utilizing physical CDs or records to digital models where users stream or acquire music from internet services, it is now easier than ever to track user preferences and this presents a new opportunity to better predict and recommend songs to users.

In building a recommendation system, various trade-offs had to be accounted for in the type of models that need to be fit. For example, some methods focus more on the metadata of the users and songs (i.e. content-based, such as the user or song characteristics - age, genre, location, etc. ) while other methods use a collaborative filtering approach where a user feedback system such as a star rating is used without considering other metadata. In this paper, the latter approach is the focus.

While songs are typically grouped into genres, it is harder to quantify exactly how similar a song is to another and what characteristics of songs a user actually prefers. Two songs by the same artist can produce very different feedback from the same user and thus, determining the exact similarity between two songs is no easy task. The benefit of collaborative filtering methods is that we assume that the tastes of a user can be estimated to be similar to another user and thus we can recommend the songs that the second user has enjoyed to the first user.

@@

\subsection{Testing}
We benchmarked our algorithms using a mean average precision score truncated at 500 (MAP@500) for each algorithm. This choice of benchmark was mainly due to MAP@500 also being used as the main method to evaluate algorithms submitted for the Kaggle MSD challenge \citep{McFee:2012:MSD:2187980.2188222}.

Specifically, MAP@500 scores take into account the first 500 recommendations given to each user. Generally, the score looks at the precision for a list of 500 song recommendations for a user and calculates the amount of correct recommendations (true positives) against the total amount of recommended songs. This was done by calculating the average precision at each element in the list of recommended items (i.e. percentage of correct items among first $k$ recommendations) and averaging these for the first 500 recommendations (i.e. for each $k \in \{1, ..., 500\}$, calculate the precision and average them together). Then, the average score is simply calculated by taking the average score over all users.

Intuitively, the score looks at the percentage of recommendations that were in-line with what the user actually liked using the testing set that were not used by the algorithms during training. This score also takes into account the order of the songs recommended so it is desirable to have the song with the highest confidence to be liked at the beginning of the recommendation list. For example, if a user had a test set list of five songs, recommending these five songs at the very beginning would give a MAP score of 100 while recommending them at the very end of the 500 songs in reverse order would only produce a MAP score of 0.0060. The MAP score can be said to be similar to the AUC of the ROC curve discussed in class without taking into account the order of recommendations.

Two different baseline benchmarks based on \citet{McFee:2012:MSD:2187980.2188222} were used. The first was to simply to recommend the top 500 most popular songs in the training dataset to every user with no user personalization. This naive method produced a MAP score of 0.0138 if the top 500 songs were based on the numbers of users who listened to the song or 0.0126 if the top 500 songs were based on the number of total plays the song had. The second baseline used had the bare amount of user personalization where the most popular songs from the artists that the user had also listened to were ordered first. Although this method was also relatively simple and naive, it resulted in a very significant improvement of the MAP score to produce a value of 0.0448.
